---
title: "Australian House Price Index"
author: "Shaun"
date: "July 29, 2017"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressMessages(library(plotly))
suppressMessages(library(xtable))
suppressMessages(library(leaflet))
suppressMessages(library(RJSONIO))
house_price <- readRDS("../shiny/cache/house.RDS")
```

## Introduction

The Australian housing market has become a contentious issue between those who already have a house and those looking to enter the market for the first time.

This application helps to identify australian housing market segments that have performed well and poorly in relation to one another in order to 
empower the user to make an informed decision.

https://swest.shinyapps.io/australian_house_price_index/

## Data Summary

- The data was obtained from <http://www.abs.gov.au/ausstats/abs@.nsf/0/510D8915596EEFE9CA257F1B001B0107?Opendocument>(Australian Bureau of Statistics)
- The data range is from `r min(house_price$date)` to `r max(house_price$date)`
- There is a total of `r dim(house_price)[1]` rows of data

## House Price Index of Residential Properties By Region

The graph below is a subet of data available on the Shiny App.
```{r echo = T, warning = F}
plot_ly(house_price[which(house_price$Measure == "Index Number" & house_price$Property.type == "Residential property"), ], x = ~date) %>%
  group_by(Region) %>%
  add_lines(y = ~Value, color = ~Region) %>%
  layout(
        title = "Australian House Price Index",
        xaxis = list(
          rangeslider = list(type = "date")
         ),
        yaxis = list(title = "Value")
    )
```


## Closing

Thank you for taking the time to view this presentation!

```{r echo=FALSE}
geocodeAdddress <- function(address) {
      url <- "http://maps.google.com/maps/api/geocode/json?address="
      url <- URLencode(paste(url, address, "&sensor=false", sep = ""))
      x <- fromJSON(url, simplify = FALSE)
      if (x$status == "OK") {
        out <- c(x$results[[1]]$geometry$location$lng,
                 x$results[[1]]$geometry$location$lat)
      } else {
        out <- NA
      }
      Sys.sleep(0.2)  # API only allows 5 requests per second
      out
}

#Obtain the Geolocations of each town
if (file.exists("./data/cities.RDS")) {
    cities <- readRDS("./data/cities.RDS")
} else {
    cities <- data.frame(city = c("Perth", "Sydney", "Australian Capital Territory", "Adelaide", "Brisbane", "Darwin","Hobart", "Melbourne"))
    count <- 1
    for (i in cities$city) {
        latlong <- geocodeAdddress(paste0("Australia, ", i))
        cities[count, "longitute"] <- latlong[1]
        cities[count, "latitude"] <- latlong[2]
        count <- count + 1
    }
    saveRDS(cities,"./data/cities.RDS")
}

 leaflet(cities) %>% 
  addTiles() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addMarkers(lng = ~longitute, 
             lat= ~latitude, 
             popup = ~paste0(
                  "City: ", city, "</br>"
             ),
            label = ~paste0("City: ", city)
  )
```

